---
title: 'APIs'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{APIs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>'
)
```

```{r setup, echo=FALSE, include=FALSE, eval=TRUE}
# Attach package
library(wildrtrax)
library(dplyr)

load("package.RData")
#save.image("package.RData")
```

## Set up WildTrax credentials

To obtain an Auth0 token, you must login into WildTrax using Auth0 (Google authorization is currently not supported). You also need to store your WildTrax user name and password as environment variables.

```{r echo=TRUE, eval=TRUE}
# Note that you need to use 'WT_USERNAME' and 'WT_PASSWORD'
Sys.setenv(WT_USERNAME = 'guest', WT_PASSWORD = 'Apple123')

```

## Authenticate

Next, you use the `wt_auth()` function to authenticate.

```{r echo=TRUE, eval=TRUE}
# Authenticate
wt_auth()

```

The Auth0 token you obtained will last for 12 hours. After that time, you will need to re-authenticate.

## Making API calls

Once authenticated, you can now use various functions that call upon the WildTrax APIs. For instance, you can use `wt_get_projects()` to see basic metadata about projects that you can download data for.  

```{r echo=TRUE, eval=TRUE, wanring=F, message=F}
# Download the project summary you have access to
my_projects <- wt_get_projects(sensor = 'ARU')

head(my_projects)
```

Using the project_id number in the download summary you can then use `wt_download_report()` to access the species data. You can also find the project_id number in the url of a WildTrax project, e.g. https://portal.wildtrax.ca/home/aru-tasks.html?projectId=605&sensorId=ARU.

```{r echo=TRUE, include=TRUE, eval=F, warning = FALSE, message = FALSE}
# Download the project report
my_report <- wt_download_report(project_id = 620, sensor_id = 'ARU', reports = "main", weather_cols = F) %>%
  tibble::as_tibble()
```
```{r echo=F, include=F, eval=T, warning = FALSE, message = FALSE}
# Download the project report
my_report <- wt_download_report(project_id = 620, sensor_id = 'ARU', reports = "main", weather_cols = F) %>%
  tibble::as_tibble()
```
```{r, eval=T}
head(my_report)

```

An easy way to download multiple projects at once is to use `wt_get_projects()` and then filter by a substring in order to get the project ids to download the data.

```{r echo=T, include=T, eval=FALSE, warning = F, message = F}
# Download all of the published Ecosystem Health ARU data to a single object
wt_get_projects(sensor_id = "ARU") %>%
  tibble::as_tibble() %>%
  dplyr::filter(grepl('^Ecosystem Health',project)) %>%
  dplyr::mutate(data = purrr::map(.x = project_id, .f = ~wt_download_report(project_id = .x, sensor_id = "ARU", weather_cols = F, reports = "main")))

```
```{r echo=T, include=F, eval=F, warning = F, message = F}
# Download all of the published Ecosystem Health ARU data to a single object
wt_get_projects(sensor_id = "ARU") %>%
  tibble::as_tibble() %>%
  dplyr::filter(grepl('^Ecosystem Health',project)) %>%
  dplyr::mutate(data = purrr::map(.x = project_id, .f = ~wt_download_report(project_id = .x, sensor_id = "ARU", weather_cols = F, reports = "main")))

```

WildTrax also pre-formats ARU to point count (PC) data depending on the type of analysis you wish to perform. See the [Boreal Avian Modelling project](https://borealbirds.ca/) website and GitHub [repositories](https://github.com/borealbirds) to find out more on integration of avian point count and ARU data.

```{r, echo=T, include=T, eval=T, warning=F, message = FALSE}
# As ARU format
my_report
```
```{r, echo=F, include=F, eval=T, warning=F, message = FALSE}
# As point count format
aru_as_pc <- wt_download_report(project_id = 620, sensor_id = 'PC', reports = "main", weather_cols = F)
```
```{r, eval=T}
# As point count format
head(aru_as_pc)

```

## Species

Downloading the WildTrax species table with `wt_get_species()` also grants you access to other valuable columns or provides a complete list of the species currently supported by WildTrax.

```{r, echo=T, include=T, eval=F, warning=F, message = FALSE}
# Download the WildTrax species table
spp_table <- wt_get_species()
```
```{r, echo=F, include=F, eval=T, warning=F, message = FALSE}
# Download the WildTrax species table
spp_table <- wt_get_species()
```
```{r, eval=T}
spp_table |> arrange(species_code)

```

You can also determine which species are allowed in and out of projects via the `wt_get_project_species()` function. 

```{r, echo=T, include=T, eval=F, warning=F, message = FALSE}
my_project_species <- wt_get_project_species(620)

my_project_species |>
  filter(species_allowed_in_project == TRUE)
```

## Data Discover

Explore species and data within WildTrax's [Data Discover](https://discover.wildtrax.ca/) by employing the `wt_dd_summary()` function. Access a portion of data, even without user privileges. Utilize `wt_auth()` to uncover data pertinent to your account or those publicly available on WildTrax.

```{r, echo=T, include=T, eval=F, warning=F, message = FALSE}
discover <- wt_dd_summary(sensor = "ARU", species = "White-throated Sparrow", boundary = NULL)
```
```{r, echo=F, include=F, eval=F, warning=F, message = FALSE}
discover <- wt_dd_summary(sensor = "ARU", species = "White-throated Sparrow", boundary = NULL)
```
```{r, eval=T}
head(discover)

```

Use custom bounding areas:

```{r, echo=T, include=T, eval=F, warning=F, message = FALSE}
# Define a polygon
my_aoi <- list(
  c(-113.96068, 56.23817),
  c(-117.06285, 54.87577),
  c(-112.88035, 54.90431),
  c(-113.96068, 56.23817)
)

discover <- wt_dd_summary(sensor = "ARU", species = "White-throated Sparrow", boundary = my_aoi)

head(discover)

```

```{r, echo=T, include=T, eval=F, warning=F, message = FALSE}
library(sf)
# Alberta bounding box
abbox <- read_sf("...shp") |> # Shapefile of Alberta
  filter(Province == "Alberta") |>
  st_transform(crs = 4326) |> 
  st_bbox()

discover <- wt_dd_summary(sensor = "ARU", species = "White-throated Sparrow", boundary = abbox)

head(discover)

```

## Sync columns

`wt_get_sync()` is the primary function used to access WildTrax **sync** APIs, such as getting locations, visits, equipment, project tasks and tags and more. 
```{r, echo=T, include=T, eval=F, warning=F, message = FALSE}
my_project_tasks <- wt_get_projects('ARU') |>
  filter(grepl('Cypress', project)) |>
  select(project_id) |>
  pull() |>
  (\(proj_id) wt_get_sync("download-tasks-by-project-id", option = "data", project = proj_id))()

my_project_tasks

```
  
