---
title: "Ultrasonic workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ultrasonic workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

from [Songscope](https://www.wildlifeacoustics.com/resources/faqs/song-scope-is-no-longer-supported) and [Kaleidoscope](https://www.wildlifeacoustics.com/products/kaleidoscope-pro) , \code`wt_songscope_tags()` and \code`wt_kaleidoscope_tags()`

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

Sys.setenv(WT_USERNAME = "guest", WT_PASSWORD = "Apple123")

```

The function `wt_songscope_tags()` reformats the output obtained from a Wildlife Acoustics Songscope recognizer. This transformation involves converting the recognizer tags into tags that do not have a method type. This makes it possible to upload each hit as a tag in a task. Similarly, the function `wt_kaleidoscope_tags()` performs the same reformatting process, but with Kaleidoscope instead. It is worth noting that this function targeted for sonic and ultrasonic species upload.

```{r eval=F, message=FALSE, warning=FALSE, include=T}
# Convert Songscope output into WildTrax tags
wt_songscope_tags(
  input, 
  output = c("env", "csv"),
  my_output_file = NULL,
  species_code,
  vocalization_type,
  score_filter,
  method = c("USPM", "1SPT"),
  task_length
)

# Convert Kaleidoscope output into WildTrax tags
wt_kaleidoscope_tags(
  input, 
  output,
  tz, 
  freq_bump = T) # Add a frequency buffer to the tag, e.g. 20000 kHz

```

Similarly you can also use GUANO metadata

```{r eval=F, message=FALSE, warning=FALSE, include=T}

wt_guano_tags()

```

You can add GRTS IDs to your location names

```{r eval=F, message=FALSE, warning=FALSE, include=T}

bats <- wt_download_report(685, 'ARU', 'location')

grts <- wt_add_grts(bats, group_locations_in_cell = TRUE)

grts


```

